---
title: "Protein Neighborhood Analysis"
author: "Maximilian Böhm"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Protein Neighborhood Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `proteinNeighbours` package provides tools for analyzing the genomic environment of proteins. Understanding the genomic context of a protein can provide valuable insights into its function, as functionally related genes are often located close to each other on the genome. This package helps you:

1. Identify neighboring genes within a specified distance
2. Annotate proteins using eggNOG-mapper with COG classification
3. Analyze the distribution of neighbor types across different protein classes
4. Visualize genomic neighborhoods and their patterns

This vignette walks you through a complete analysis workflow, from data preparation to result interpretation.

# Installation

## Dependencies

The package has several dependencies that need to be installed:

### R Packages

These will be installed automatically when you install the package:

```{r dependencies, eval = FALSE}
install.packages(c("tidyverse", "ape", "dplyr", "gggenes", "ggplot2", 
                  "RColorBrewer", "yaml", "purrr", "readr", "stringr", 
                  "tidyr", "withr", "reshape2"))

# Optional but recommended packages
install.packages(c("rmarkdown", "knitr", "logger", "DT", "plotly"))
```

### External Tools

The package relies on two external tools:

1. **NCBI Entrez Direct utilities** - For retrieving sequences from NCBI
   Installation instructions: https://www.ncbi.nlm.nih.gov/books/NBK179288/

2. **eggNOG-mapper** - For functional annotation of proteins
   Installation instructions: http://eggnog-mapper.embl.de/

Install eggNOG-mapper with:

```bash
pip install eggnog-mapper
```

# Getting Started

## Data Preparation

Before running an analysis, you need to prepare several input files.
These files are generated when following the overall pipe line of repositories illustrated bellow:
1. protein-per-organism
2. protein-to-genome
3. protein-neighbours

The files need to be collected and placed in a specific directory structure. The default data directory is `data/`, but you can change this in the configuration file.

### Required Files

Place these files in your data directory (default is `data/`):

1. **proteins.csv** - A file containing protein accession numbers (one per line)
2. **assm_accs.csv** - A file containing genome assembly accession numbers (one per line)
3. **assm_accs_protein.csv** - A mapping between proteins and assemblies (two columns: assembly, protein)

### GFF Files

The package needs GFF files for the assemblies. These should be placed in:

```
data/ncbi_dataset/data/{assembly_id}/genomic.gff
```

You can download these files from NCBI using the datasets command-line tool:

```bash
datasets download genome accession --inputfile assm_accs.csv --include gff3
unzip ncbi_dataset.zip -d data/
```
Alternatively, use my pipe line to download the GFF files automatically. (protein-to-genome repository)
### Optional Files

For advanced analyses, you can create these additional files:

#### Clade Files

Create a `clades` folder in your data directory with files for each clade:

```
data/clades/CladeA.txt
data/clades/CladeB.txt
...
```

Each file should contain protein IDs belonging to that clade.

#### Representative Files

For protein alias mapping (useful when protein accession numbers change):

```
data/representatives/ipg_representative.txt
data/representatives/pdb_representative.txt
data/representatives/cluster_representative.txt
```

## Basic Usage

Here's a simple example of running an analysis:

```{r basic-usage, eval = FALSE}
source(main.R)

# Run with default settings
results <- main()

# Run with custom configuration file
results <- main("config/my_config.yaml")

# Override specific parameters
results <- main(override_params = list(
  "analysis.basepairs" = 500,
  "analysis.max_neighbors" = 20,
  "annotation.tool" = "eggnog"
))
```

# Configuration

## Configuration File

The package uses a YAML configuration file for all parameters. The default location is `config/config.yaml`, but you can specify a different file.

Here's an example of a configuration file:

```yaml
# Input/Output settings
paths:
  base_dir: "data"
  output_dir: "output"
  ncbi_dataset: "data/ncbi_dataset"

# Input files
files:
  proteins: "proteins.csv"
  assemblies: "assm_accs.csv"
  protein_assembly: "assm_accs_protein.csv"

# Analysis parameters
analysis:
  basepairs: 300
  max_neighbors: 15
  date: null  # Uses current date if null

# Annotation settings
annotation:
  tool: "eggnog"
  eggnog:
    db_dir: "data/eggnog_db"
    cpu: 4
```

## Parameter Overriding

You can override configuration parameters without modifying the file:

```{r override-params, eval = FALSE}
# Override specific parameters
results <- main(override_params = list(
  "analysis.basepairs" = 500,
  "paths.base_dir" = "my_data",
  "annotation.eggnog.cpu" = 8
))
```

# Workflow in Detail

Let's walk through a complete analysis workflow:

## 1. Loading Data

The first step is to load protein and assembly data:

```{r load-data, eval = FALSE}
# Read protein and assembly data
protein_assembly_data <- read_protein_assembly_data(
  protein_file = "proteins.csv",
  assembly_file = "assm_accs.csv",
  protein_assembly_file = "assm_accs_protein.csv",
  PATH = "data"
)

# Get protein aliases
protein_aliases <- read_representatives(PATH = "data")
```

## 2. Identifying Neighbors

Next, identify neighboring proteins:

```{r neighbors, eval = FALSE}
# Find neighbors for all proteins
all_neighbours <- collect_all_neighbours(
  protein_assembly_data$protein_assembly,
  basepairs = 300,
  max_neighbors = 15,
  PATH = "data"
)

# Get information about all proteins
all_protein <- collect_all_protein_info(
  protein_assembly_data$protein_assembly,
  PATH = "data"
)

# Combine all data
all_data <- rbind(all_neighbours, all_protein)
```

## 3. Protein Annotation

Annotate proteins with eggNOG-mapper:

```{r annotation, eval = FALSE}
# Annotate proteins
annotation_results <- analyze_proteins(
  df = all_neighbours,
  column = 'ID',
  config = config
)

# Analyze neighbor types
amount_of_neighbours(annotation_results)
```

## 4. Combining Data

Combine all data for analysis:

```{r combine-data, eval = FALSE}
# Get clade information
clades <- read_clades(PATH = "data")

# Check for manual annotations
annotated_neighbours <- read_annotations(current_date)

# Combine all data
combined_df <- combine_and_plot(
  neighbours_data = all_data,
  cog_data = annotation_results,
  clade_assign = clades,
  neighbour_annotations = annotated_neighbours
)
```

## 5. Visualization

Create visualizations:

```{r visualization, eval = FALSE}
# Plot distribution of neighbors by clade
plot_neighbours_per_clade(combined_df)

# Plot distribution excluding unknowns
plot_neighbours_per_clade(
  combined_df,
  exclude_unknown_clade = TRUE,
  exclude_unknown_cog = TRUE
)

# Create correlation matrix
make_correlation_matrix(
  combined_df %>%
    select(PIGI, assembly, clade) %>%
    unique() %>%
    select(assembly, clade),
  unique(combined_df$clade)
)

# Create histograms
create_clade_histograms2(
  combined_df %>%
    select(PIGI, assembly, clade) %>%
    unique()
)
```

## 6. Manual Annotation

After running the initial analysis, you can manually annotate neighbor types:

1. Examine `output/{date}/types_of_neighbours.csv`
2. Create a new file named `types_of_neighbours_annotated.csv` in the same directory
3. Add a fourth column with your annotations
4. Re-run the analysis to incorporate your annotations

```{r manual-annotation, eval = FALSE}
# Re-run with manual annotations
annotated_neighbours <- read_annotations(current_date)

combined_df <- combine_and_plot(
  neighbours_data = all_data,
  cog_data = annotation_results,
  clade_assign = clades,
  neighbour_annotations = annotated_neighbours
)

# Plot with annotations
plot_neighbours_per_clade(
  combined_df %>% mutate(COG_LETTER = ANNOTATION),
  exclude_unknown_clade = TRUE,
  exclude_unknown_cog = TRUE,
  output_path = "annotated_neighbours"
)
```

# Advanced Topics

## Working with Large Datasets

For large datasets, here are some optimization strategies:

```{r large-data, eval = FALSE}
# Use parallel processing for eggNOG-mapper
results <- main(override_params = list(
  "annotation.eggnog.cpu" = 8  # Use 8 CPU cores
))

# Load existing results instead of recalculating
results <- main(override_params = list(
  "analysis.date" = "2025-02-20"  # Load results from a specific date
))
```

## Custom Visualization

You can create custom visualizations:

```{r custom-vis, eval = FALSE}
# Plot neighbors for a specific protein
plot_neighbours(all_neighbours, "YP_123456.1", width = 25, height = 7)

# Create a custom correlation plot
correlation_matrix <- calculate_correlation(df, vector)
custom_plot <- plot_correlation_matrix(correlation_matrix) +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("My Custom Correlation Plot")
```

## Integrating with Other Analyses

You can integrate the results with other analyses:

```{r integration, eval = FALSE}
# Extract COG distribution
cog_distribution <- combined_df %>%
  filter(is.neighbour) %>%
  count(COG_LETTER) %>%
  arrange(desc(n))

# Export for use in other tools
write.csv(cog_distribution, "cog_distribution.csv", row.names = FALSE)

# Generate a network graph (requires the igraph package)
library(igraph)
network_data <- combined_df %>%
  filter(is.neighbour) %>%
  select(PIGI, ID, COG_LETTER) %>%
  rename(from = PIGI, to = ID, type = COG_LETTER)

g <- graph_from_data_frame(network_data, directed = FALSE)
plot(g, vertex.color = "lightblue", vertex.size = 5)
```

# Troubleshooting

## Common Issues

### Missing GFF Files

If you encounter errors about missing GFF files:

```{r missing-gff, eval = FALSE}
# Check GFF file paths
gff_paths <- file.path("data/ncbi_dataset/data", 
                      protein_assembly_data$protein_assembly[, 1], 
                      "genomic.gff")
missing_files <- gff_paths[!file.exists(gff_paths)]
cat("Missing GFF files:", length(missing_files), "\n")
```

### eggNOG-mapper Errors

If eggNOG-mapper fails:

```bash
# Check eggNOG-mapper installation
emapper.py --version

# Download eggNOG database if needed
download_eggnog_data.py -y
```

### Out of Memory

For large datasets, you may encounter memory issues:

```{r memory-issue, eval = FALSE}
# Process proteins in batches
protein_batches <- split(protein_assembly_data$protein_assembly, 
                        ceiling(seq_along(protein_assembly_data$protein_assembly[,1]) / 100))

all_neighbours <- data.frame()
for (batch in protein_batches) {
  batch_neighbours <- collect_all_neighbours(batch, basepairs = 300, max_neighbors = 15)
  all_neighbours <- rbind(all_neighbours, batch_neighbours)
  gc()  # Force garbage collection
}
```

# Conclusion

The `proteinNeighbours` package provides a comprehensive toolkit for analyzing protein genomic neighborhoods. By understanding the genomic context of proteins, you can gain valuable insights into protein function, evolution, and the organization of genomic regions.

For further assistance, please refer to the function documentation or file an issue on GitHub.

# Session Info

```{r session-info}
sessionInfo()
```
