---
title: "Protein Neighborhood Analysis Report"
author: "Protein Neighbours Package"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
params:
  combined_df: NULL
  config: NULL
  output_dir: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(DT)
library(ggplot2)
library(plotly)

# Get parameters
combined_df <- params$combined_df
config <- params$config
output_dir <- params$output_dir
```

# Analysis Overview

This report summarizes the protein neighborhood analysis performed on `r nrow(params$combined_df %>% filter(!is.neighbour))` proteins, identifying `r nrow(params$combined_df %>% filter(is.neighbour))` neighboring proteins.

## Configuration Summary

```{r config-summary}
# Convert config to a simplified data frame for display
config_flat <- data.frame(
  Parameter = c(
    "Base Directory",
    "Output Directory",
    "Base Pairs",
    "Max Neighbors",
    "Annotation Tool",
    "Date",
    "Exclude Unknown Clades",
    "Exclude Unknown COGs"
  ),
  Value = c(
    config$paths$base_dir,
    config$paths$output_dir,
    config$analysis$basepairs,
    config$analysis$max_neighbors,
    config$annotation$tool,
    config$analysis$date,
    ifelse(config$visualization$exclude_unknown_clade, "Yes", "No"),
    ifelse(config$visualization$exclude_unknown_cog, "Yes", "No")
  ),
  stringsAsFactors = FALSE
)

kable(config_flat, caption = "Analysis Configuration")
```

## Input Files

```{r input-files}
# Read input file info if it exists
input_file_info_path <- file.path(output_dir, "input_file_info.csv")
if (file.exists(input_file_info_path)) {
  input_file_info <- read.csv(input_file_info_path)
  kable(input_file_info, caption = "Input File Information")
} else {
  cat("Input file information not available.")
}
```

# Analysis Results

## Protein and Neighbor Summary

```{r protein-summary}
# Get summary statistics
protein_count <- nrow(combined_df %>% filter(!is.neighbour) %>% distinct(PIGI))
neighbor_count <- nrow(combined_df %>% filter(is.neighbour))
clade_count <- length(unique(combined_df$clade))
assembly_count <- length(unique(combined_df$assembly))

summary_df <- data.frame(
  Metric = c("Total Proteins", "Neighbors", "Clades", "Assemblies"),
  Count = c(protein_count, neighbor_count, clade_count, assembly_count),
  stringsAsFactors = FALSE
)

kable(summary_df, caption = "Summary Statistics")
```

## Clade Distribution

```{r clade-distribution}
# Create a clade distribution plot
clade_dist <- combined_df %>%
  filter(!is.neighbour) %>%
  count(clade) %>%
  arrange(desc(n)) %>%
  mutate(clade = factor(clade, levels = clade))

ggplot(clade_dist, aes(x = reorder(clade, n), y = n, fill = clade)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Clade", y = "Count", title = "Distribution of Proteins by Clade") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Neighbor Type Distribution

```{r neighbor-type-distribution}
# Create a neighbor type distribution plot
if ("COG_LETTER" %in% colnames(combined_df)) {
  neighbor_type_dist <- combined_df %>%
    filter(is.neighbour) %>%
    count(COG_LETTER) %>%
    arrange(desc(n)) %>%
    mutate(COG_LETTER = factor(COG_LETTER, levels = COG_LETTER))
  
  ggplot(neighbor_type_dist, aes(x = reorder(COG_LETTER, n), y = n, fill = COG_LETTER)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "COG Letter", y = "Count", title = "Distribution of Neighbors by COG Type") +
    theme_minimal() +
    theme(legend.position = "none")
} else {
  cat("COG annotation information not available.")
}
```

## Top Proteins by Neighbor Count

```{r top-proteins}
# Show proteins with the most neighbors
top_proteins <- combined_df %>%
  filter(is.neighbour) %>%
  count(PIGI) %>%
  arrange(desc(n)) %>%
  head(20)

kable(top_proteins, caption = "Top 20 Proteins by Neighbor Count")
```

## Annotation Results

```{r annotation-results}
# Show annotation summary
if ("COG_LETTER" %in% colnames(combined_df) && "COG_NAME" %in% colnames(combined_df)) {
  annotation_summary <- combined_df %>%
    filter(is.neighbour) %>%
    count(COG_LETTER, COG_NAME) %>%
    arrange(desc(n)) %>%
    head(20)
  
  kable(annotation_summary, caption = "Top 20 Annotation Categories")
} else {
  cat("Annotation category information not available.")
}
```

## Sample Proteins and Their Neighbors

```{r sample-proteins}
# Display a few sample proteins and their neighbors
if (nrow(combined_df) > 0) {
  sample_proteins <- combined_df %>%
    filter(!is.neighbour) %>%
    distinct(PIGI) %>%
    sample_n(min(5, nrow(.)))
  
  sample_data <- combined_df %>%
    filter(PIGI %in% sample_proteins$PIGI)
  
  DT::datatable(
    sample_data %>% select(PIGI, ID, clade, assembly, is.neighbour),
    caption = "Sample Proteins and Their Neighbors",
    options = list(pageLength = 10)
  )
} else {
  cat("No protein data available for display.")
}
```

# Generated Plots

The following plots were generated during the analysis and are available in the output directory:

```{r list-plots, results='asis'}
# List plot files
plot_files <- list.files(output_dir, pattern = "\\.(png|svg)$", full.names = TRUE)

if (length(plot_files) > 0) {
  cat("## Available Plots\n\n")
  
  for (plot_file in plot_files) {
    plot_name <- basename(plot_file)
    cat(sprintf("- %s\n", plot_name))
  }
  
  # Display a few example plots
  png_files <- list.files(output_dir, pattern = "\\.png$", full.names = TRUE)
  
  if (length(png_files) > 0) {
    cat("\n## Example Visualizations\n\n")
    
    for (i in 1:min(3, length(png_files))) {
      cat(sprintf("### %s\n\n", basename(png_files[i])))
      cat(sprintf("![%s](%s)\n\n", basename(png_files[i]), png_files[i]))
    }
  }
} else {
  cat("No plots were generated during the analysis.")
}
```

# Conclusion and Next Steps

This report provides an overview of the protein neighborhood analysis results. For detailed examination, please refer to the generated CSV files and plots in the output directory.

## Recommended Next Steps

1. Examine the detailed plots for specific patterns of interest
2. Review the annotation results for functional insights
3. Consider running the analysis with different parameters for sensitivity analysis
4. Perform more detailed analysis on specific clades or protein groups of interest

---

*Report generated on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`*
